Load Shifting Versus Manual Frequency Reserve: \\ Which One is More Appealing to Flexible Loads in Denmark?

### 2024-01-12 tex/paper_flex_v2.tex
Credit author statement
Submission

\begin{document}
\title{
    %A Pioneering Market Framework for Promoting Demand-side Flexibility in Denmark 
    %A New Ecosystem for Demand-side Flexibility Aggregators in Denmark 
    Load Shifting Versus Manual Frequency Reserve: \\ Which One is More Appealing to Flexible Loads?
}
% Pioneering Market Framework for the Future Power System: the Case of Denmark
%\author{Peter Alexander Vistar Gade$^{\textrm{a}}$, Henrik Bindner, Jalal
%Kazempour}
\author[1,2]{Peter A.V. Gade\footnote{Corresponding author. Tel.: +45 24263865. \\ Email addresses: pega@dtu.dk (P.A.V. Gade), Trygve.Skjotskift@ibm.com (T. Skjøtskift), chazi@dtu.dk (C. Ziras), hwbi@dtu.dk (H.W. Bindner), jalal@dtu.dk (J. Kazempour).}}
\author[2]{Trygve Skjøtskift}
\author[1]{Charalampos Ziras}
\author[1]{Henrik W. Bindner}
\author[1]{Jalal Kazempour}
\affil[1]{Department of Wind and Energy Systems, Technical University of Denmark, Kgs. Lyngby, Denmark}
\affil[2]{IBM Client Innovation Center, Copenhagen, Denmark}
%\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

@@

This paper investigates how a  thermostatically controlled load can deliver flexibility either in form of manual frequency restoration reserves (mFRR) or load shifting, and which one is financially more appealing to such a load. %We discuss advantages and disadvantages of the two and their appeal from a monetary point of view.
A supermarket freezer is considered as a representative flexible load, and a grey-box model  describing its temperature dynamics is developed using real data from a supermarket in Denmark. Taking into account price and activation uncertainties, a two-stage stochastic mixed-integer linear program is formulated to maximize the flexibility value from the freezer. For practical reasons, we propose a linear policy to determine regulating power bids, and then linearize the mFRR activation conditions through the  McCormick relaxation approach. For computational ease, we develop a decomposition technique, splitting the problem to a set of smaller sub-problems, one per scenario. %An out-of-sample evaluation was done on unseen Danish market price data for 2022. 
%
%We observed that load shifting was more profitable, but had a greater impact on the air and food temperatures in the freezer as opposed to mFRR that depends on the system state and bid price for activation.
%
%solution strategies for mFRR are presented: one with a simple policy and the other one with a dynamically updated policy. The proposed model accurately captures the timeline of decisions for bidding in manual frequency reserves.
%by using McCormick relaxation.
Examined on an out-of-sample simulation based on real Danish spot and balancing market prices in 2022, load shifting shows to be more profitable than mFRR provision, but is also more consequential for temperature deviations in the freezer.

@@

Power consumers are looking to reduce costs by any means due to sky-rocketing energy and gas prices. In Denmark, supermarkets are especially exposed due to their high energy consumption, and they are actively looking into initiatives to reduce their electricity bill. One of these initiatives is the application of demand-side flexibility, whereby a supermarket can shift consumption in time when spot (day-ahead) prices are comparatively lower or when the power system needs up-regulation services. There are thousands of supermarkets in Denmark willing to provide flexibility, and their freezers are especially suited for this as they constitute a big share of their energy consumption.

IBM is currently developing a software platform, called \textit{Flex Platform}, aiming to harness flexibility from industrial and commercial consumers for bidding in ancillary service markets via a Balance Responsible Party (BRP). One large supermarket chain in Denmark has already signed up for the \textit{Flex Platform}, and another supermarket chain is currently exploring the possibility of joining.
While these supermarket chains are keen on the provision of flexibility, the estimation of their monetary benefit still remains an open question. We aim to address it by considering two potential ways for supermarkets to exploit their consumption flexibility: (\textit{i}) load shifting, and (\textit{ii}) the provision of manual Frequency Restoration Reserve (mFRR)
%\footnote{The term will be used for the remainder of the paper and is equivalent to tertiary reserves.})
services. We  develop  a stochastic optimization tool capturing the temperature dynamics and therefore consumption flexibility of supermarkets, and compute the monetary benefit of flexibility implemented via load shifting or mFRR service provision. We use two scenario generation  strategies: one using a five-day lookback on spot and balancing market prices in Denmark, and the other one using those price data in 2021. We then systematically compare these two strategies ex-post via an out-of-sample simulation based on price data in 2022.

\subsection{Dilemma: Load Shifting Versus mFRR}
%
For power consumers who are actively looking into cost saving solutions, it is important to estimate the monetary benefit of flexibility. In Nord Pool, spot prices are announced at 2pm for the next day, so supermarkets who purchase power through a retailer at  spot prices, have a chance to shift their consumption plan for the next day, accordingly. From their perspective, it is natural to first look into load shifting as the most obvious and straightforward way to utilize flexibility. Here, load shifting simply refers to shifting consumption to another point in time, typically when spot prices are lower.

@@

For simplicity, we assume that one freezer delivering mFRR can be scaled to many freezers in the same way. Therefore,
%(and likewise for load shifting).
demand response for consumers participating in the mFRR market versus load shifting will  look similar. From the perspective of consumers, it is of great interest to explore the monetary benefit of mFRR versus load shifting. Similarly, it is important for TSOs to understand the incentive of consumers to provide mFRR services opposed to simply shifting load.

\subsection{Research questions and literature review}
%
This work aims to investigate and answer the following research questions: (\textit{i}) how can the flexibility of a supermarket freezer be characterized? (\textit{ii}) how much monetary benefit can be earned by the mFRR provision opposed to load shifting? and finally (\textit{iii})
what are the pros and cons of participating in the mFRR market versus load shifting? Some of these questions have been  looked at  in the literature, so we provide a  literature review to pinpoint where new work is required, and how our work differs from the current literature.

@@

In \cite{petersen2012eso2} and \cite{pedersen2013direct}, a complete white-box model of a supermarket refrigeration system is presented and validated against real-life data. It is also shown how such a system can provide demand response. This work serves as a  benchmark for any grey-box model of a supermarket refrigeration system. However, such an approach is hard to scale and requires complete knowledge of every refrigeration system. In \cite{pedersen2016improving}, a second-order model is used to model the food and air temperature in a freezer in an experimental setting. It is shown how food temperature has much slower dynamics than air temperature. In \cite{hao2014aggregate}, a simple first-order virtual battery model for a Thermostatically Controlled Load (TCL) is presented. This model also constitutes the starting point for modeling temperature dynamics in this work, as it has an intuitive interpretation. A similar bucket model is introduced in \cite{petersen2013taxonomy}.

@@

\subsection{Our contribution and outline}
%
To the best of our knowledge, no work has investigated the monetary value and incentives for providing mFRR services versus load shifting, although both have been studied extensively. This work aims to fill that gap by taking a holistic view on the incentives from the flexible consumer's perspective. Furthermore, when stating the objective functions for mFRR, most studies have assumed a simplified market structure, and neglect the fact that the aggregator and the consumer are not necessarily the same entity. We discuss the consequences of this assumption in relation to the market structure in Denmark \cite{gade2022ecosystem}, and in the context of mFRR and load shifting.

@@

The rest of the paper is organized as follows. Section \ref{sec:monetizing_flex} describes modeling a supermarket freezer as a TCL. Thereafter, mFRR and load shifting are described in detail and their objective functions are stated. Section \ref{sec:OptimizationModel} presents the proposed stochastic optimization problem.
%Furthermore, the two solution strategies for solving the optimization problem for mFRR are introduced. 
Section \ref{sec:results} provides  results for a case study. Section \ref{sec:conclusion} concludes the paper. Finally, an Appendix provides the full model formulation.

\section{Describing flexibility of a TCL}\label{sec:monetizing_flex}

There are several ways to monetize flexibility from TCLs. In this section, we focus on mFRR and load shifting. First, we describe how to mathematically model a TCL as a flexible resource. Second, we describe how to monetize the flexibility from TCLs for mFRR and load shifting, and provide the objective functions in both cases. For mFRR, the objective function includes all costs and revenues for the BRP, while for load shifting, the objective function only includes the flexible demand's perspective. This approach explicitly shows the situation where flexible demand is activated without including the BRP, as this is more realistic.

@@

\section{Numerical Results and Discussion}\label{sec:results}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figures/spot_single_case.png}
    \caption{Example of load shifting in a representative day (an in-sample scenario). \textbf{Top}: Baseline consumption of the freezer and the power profile after load shifting. \textbf{Middle}: Air and food temperature dynamics. \textbf{Bottom}: Spot market prices.}
    \label{fig:fig_first_case}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figures/mFRR_single_case.png}
    \caption{Example of mFRR  provision  in a representative day (an in-sample scenario). \textbf{Top}: Reservation capacities and baseline power of the freezer. \textbf{Upper middle}: Air and food temperature dynamics. \textbf{Lower middle}: mFRR activation in this specific scenario, i.e., when $\lambda_{h}^{\rm{bid}} \leq \lambda_{h}^{\rm{b}}-\lambda_{h}^{\rm{s}}$, $\lambda_{h}^{\rm{b}} > \lambda_{h}^{\rm{s}}$, and $p^{\rm{r},\uparrow}_{h} > 0$. \textbf{Bottom}: Spot and balancing market price as well as regulating power bids in this specific scenario.}
    \label{fig:fig_second_case}
\end{figure}

@@

@@

@@

@@

\section*{Credit author statement}

>>>>

\begin{document}
\title{
    %A Pioneering Market Framework for Promoting Demand-side Flexibility in Denmark 
    %A New Ecosystem for Demand-side Flexibility Aggregators in Denmark 
    Load Shifting Versus Manual Frequency Reserve: \\ Which One is More Appealing to Flexible Loads in Denmark?
}
% Pioneering Market Framework for the Future Power System: the Case of Denmark
%\author{Peter Alexander Vistar Gade$^{\textrm{a}}$, Henrik Bindner, Jalal
%Kazempour}
\author[1,2]{Peter A.V. Gade\footnote{Corresponding author. Tel.: +45 24263865. \\ Email addresses: pega@dtu.dk (P.A.V. Gade), Trygve.Skjotskift@ibm.com (T. Skjøtskift), chazi@dtu.dk (C. Ziras), hwbi@dtu.dk (H.W. Bindner), jalal@dtu.dk (J. Kazempour).}}
\author[2]{Trygve Skjøtskift}
\author[1]{Charalampos Ziras}
\author[1]{Henrik W. Bindner}
\author[1]{Jalal Kazempour}
\affil[1]{Department of Wind and Energy Systems, Technical University of Denmark, Kgs. Lyngby, Denmark}
\affil[2]{IBM Client Innovation Center, Copenhagen, Denmark}
%\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

@@

This paper investigates %how  thermostatically controlled loads can deliver flexibility 
demand-side flexibility provision in two  forms of manual Frequency Restoration Reserve (mFRR) services and load shifting, and explores which one is financially more appealing to flexible loads in Denmark. While mFRR is an ancillary service required for the system and being bought by the system operator, load shifting is an individual act of the flexible load in response to the variation of hourly electricity prices.  %We discuss advantages and disadvantages of the two and their appeal from a monetary point of view.
Without loss of generalization, we consider a supermarket freezer  as a representative flexible load, and develop a grey-box model  describing its temperature dynamics using real data from a supermarket in Denmark. Taking into account price and activation uncertainties, a stochastic mixed-integer linear program is formulated to maximize the flexibility value from the freezer. %For practical reasons, we propose a linear policy to determine regulating power bids.
%, and then linearize the mFRR activation conditions. 
%For computational ease, we develop a decomposition technique, splitting the problem to a set of smaller sub-problems, one per scenario. %An out-of-sample evaluation was done on unseen Danish market price data for 2022. 
%
%We observed that load shifting was more profitable, but had a greater impact on the air and food temperatures in the freezer as opposed to mFRR that depends on the system state and bid price for activation.
%
%solution strategies for mFRR are presented: one with a simple policy and the other one with a dynamically updated policy. The proposed model accurately captures the timeline of decisions for bidding in manual frequency reserves.
%by using McCormick relaxation.
Examined on an ex-post simulation based on  Danish spot and balancing market prices in 2022, load shifting shows to be more profitable than mFRR provision, but is also more consequential for temperature deviations in the freezer. This indicates the need for regulatory measures by the Danish  system operator to make the provision of ancillary services more attractive to  flexible loads.

@@

Power consumers are looking to reduce costs by any means due to sky-rocketing energy and gas prices. In Denmark, supermarkets are especially exposed due to their high energy consumption, and they are actively looking into initiatives to reduce their electricity bill. One of these initiatives is the application of demand-side flexibility \cite{Iacopo,towards}, whereby a supermarket can shift consumption in time when spot (day-ahead) prices are comparatively lower or when the power system needs up-regulation services. There are thousands of supermarkets in Denmark willing to provide flexibility, and their freezers are especially suited for this as they constitute a big share of their energy consumption.

IBM is currently developing a software platform, called \textit{Flex Platform}, aiming to harness flexibility from industrial and commercial consumers for bidding in ancillary service markets via a Balance Responsible Party (BRP). One large supermarket chain in Denmark has already signed up for the \textit{Flex Platform}, and another supermarket chain is currently exploring the possibility of joining.
While these supermarket chains are keen on the provision of flexibility, the estimation of their monetary benefit still remains an open question. We aim to address it by considering two potential ways for supermarkets to exploit their consumption flexibility: (\textit{i}) load shifting, and (\textit{ii}) the provision of manual Frequency Restoration Reserve (mFRR)
%\footnote{The term will be used for the remainder of the paper and is equivalent to tertiary reserves.})
services. We  develop  a stochastic optimization tool capturing the temperature dynamics and therefore consumption flexibility of supermarkets, and compute the monetary benefit of flexibility implemented via load shifting or mFRR service provision. We use two scenario generation  strategies: one using a five-day lookback on spot and balancing market prices in Denmark, and the other one using those price data in 2021. We then systematically compare these two strategies ex-post via an out-of-sample simulation based on Danish price data in 2022.

\subsection{Dilemma: Load Shifting Versus mFRR}
%
For power consumers who are actively looking into cost saving solutions, it is important to estimate the monetary benefit of flexibility. In Nord Pool, spot prices are announced at 2pm for the next day, so supermarkets who purchase power through a retailer at  spot prices, have a chance to shift their consumption plan for the next day, accordingly. From their perspective, it is natural to first look into load shifting as the most obvious and straightforward way to utilize flexibility. Here, load shifting simply refers to shifting consumption to another point in time, typically when spot prices are lower \cite{shift}.

@@

Without loss of generalization, we assume that one freezer delivering mFRR can be scaled to many freezers in the same way. Therefore,
%(and likewise for load shifting).
demand response for consumers participating in the mFRR market versus load shifting will  look similar. From the perspective of consumers, it is of great interest to explore the monetary benefit of mFRR versus load shifting. Similarly, it is important for TSOs to understand the incentive of consumers to provide mFRR services opposed to simply shifting load in time.

\subsection{Research questions and literature review}
%
Using real data from Denmark, this work aims to investigate and answer the following three research questions: (\textit{i}) how can the flexibility of a supermarket freezer as a flexible load be characterized?
(\textit{ii}) how much monetary benefit can this flexible load earn by the mFRR provision opposed to load shifting? and finally (\textit{iii})
what are the pros and cons of participating in the Danish mFRR market versus load shifting? Unlike the second and third questions, the first one has been addressed extensively in the literature. Therefore, the following literature review focuses on the first question, where we also differentiate our work from the current literature. To the best of our knowledge, there is no  work in the existing literature addressing the second and third questions.

%we provide a  literature review to pinpoint where new work is required, and how our work differs from the current literature.

@@

In \cite{petersen2012eso2} and \cite{pedersen2013direct}, a complete white-box model of a supermarket refrigeration system is presented and validated against real-life data. It is also shown how such a system can provide demand response. This work serves as a  benchmark for any grey-box model of a supermarket refrigeration system. However, such an approach is hard to scale and requires complete knowledge of every refrigeration system. In \cite{pedersen2016improving}, a second-order model is used to model the food and air temperature in a freezer in an experimental setting. It is shown how food temperature has much slower dynamics than air temperature. In \cite{hao2014aggregate}, a simple first-order virtual battery model for a thermostatically controlled load is presented. This model also constitutes the starting point for modeling temperature dynamics in this work, as it has an intuitive interpretation. A similar bucket model is introduced in \cite{petersen2013taxonomy}.

@@

\subsection{Our contribution and outline}
%
To the best of our knowledge, no work has investigated the monetary value and incentives for providing mFRR services versus load shifting, although both have been studied extensively. This work aims to fill that gap by taking a holistic view on the incentives from the flexible consumer's perspective in Denmark. Furthermore, when stating the objective functions for mFRR, most studies have assumed a simplified market structure, and neglect the fact that the aggregator and the consumer are not necessarily the same entity. We discuss the consequences of this assumption in relation to the market structure in Denmark \cite{gade2022ecosystem}, and in the context of mFRR and load shifting.

@@

The rest of the paper is organized as follows. Section \ref{sec:monetizing_flex} describes modeling a supermarket freezer as a thermostatically controlled load. Thereafter, mFRR and load shifting are described in detail and their objective functions are stated. Section \ref{sec:OptimizationModel} presents the proposed stochastic optimization problem.
%Furthermore, the two solution strategies for solving the optimization problem for mFRR are introduced. 
Section \ref{sec:results} provides  results for a case study. Section \ref{sec:conclusion} concludes the paper. Finally, an Appendix provides the full model formulation.

\section{Describing flexibility of a thermostatically controlled load}\label{sec:monetizing_flex}

There are several ways to monetize flexibility from a Thermostatically Controlled Load (TCL). In this section, we focus on mFRR and load shifting. First, we describe how to mathematically model a TCL as a flexible resource. Second, we describe how to monetize the flexibility from TCLs for mFRR and load shifting, and provide the objective functions in both cases. For mFRR, the objective function includes all costs and revenues for the BRP, while for load shifting, the objective function only includes the flexible demand's perspective. This approach explicitly shows the situation where flexible demand is activated without including the BRP, as this is more realistic.

@@

\section{Numerical results and discussion}\label{sec:results}

@@

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\columnwidth]{figures/spot_single_case.png}
    \caption{Example of load shifting in a representative day (an in-sample scenario). \textbf{Top}: Baseline consumption of the freezer and the power profile after load shifting. \textbf{Middle}: Air and food temperature dynamics. \textbf{Bottom}: Spot market prices.}
    \label{fig:fig_first_case}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=\columnwidth]{figures/mFRR_single_case.png}
    \caption{Example of mFRR  provision  in a representative day (an in-sample scenario). \textbf{Top}: Reservation capacities and baseline power of the freezer. \textbf{Upper middle}: Air and food temperature dynamics. \textbf{Lower middle}: mFRR activation in this specific scenario, i.e., when $\lambda_{h}^{\rm{bid}} \leq \lambda_{h}^{\rm{b}}-\lambda_{h}^{\rm{s}}$, $\lambda_{h}^{\rm{b}} > \lambda_{h}^{\rm{s}}$, and $p^{\rm{r},\uparrow}_{h} > 0$. \textbf{Bottom}: Spot and balancing market price as well as regulating power bids in this specific scenario.}
    \label{fig:fig_second_case}
\end{figure}

@@

To balance the power grid and ensure its stability in the future, there is a need to integrate industrial and commercial flexible demands into existing ancillary services where mFRR is the most energy-intensive one. To that end, it is crucial to estimate and model flexibility from flexible demands, especially thermostatically controlled loads.

@@

From a policy perspective, it is concerning if load shifting happens to be more profitable than mFRR provision. It might be of great interest for TSOs to ensure flexible demands have an incentive to deliver mFRR to ensure stability in the power grid and avoid new peaks and imbalances. Such incentives could manifest in green credits for providing emission free mFRR or an add-on to the mFRR reservation price. From the perspective of the flexible demand, periods of high volatility in spot prices can readily be used for load shifting although great care should be taken with respect to temperature deviations. Provision of mFRR is more attractive with respect to temperature deviations, but it is much more complicated to get started with mFRR whereas load shifting does not depend on other market participants, hence load shifting is a simple and profitable revenue stream for flexible demands.

@@

\section*{Credit author statement}
Peter Gade conceived and designed study, developed the code, and write the initial draft of the manuscript. All authors contributed considerably to developing the idea, thoroughly discussed the model, scenarios, numerical results, and policy insights, and also revised the final manuscript.
